\Large{Learning to Pivot with Adversarial Networks}

### 2016-11-02 tex/paper.tex
small things

Assuming a gaussian prior $z \sim {\cal N}(0,1)$, we start by generating
training data $\{ x_i, y_i, z_i \}_{i=1}^N$, from which we train a neural
network classifier $f$ minimizing ${\cal L}_f(\theta_f)$ without considering its
adversary $r$. The network architecture comprises 2 dense hidden layers of 20
nodes with ReLU activations, followed  by a dense output layer with a single
node with a sigmoid activation. As shown in Fig.~\ref{fig:toy}, the resulting
classifier is not pivotal, as the conditional probability densities of its
decision scores $f(X;\theta_f)$ show large discrepancies between values $z$ of
the nuisance. While not shown here, a classifier trained only from data
generated at the nominal value $Z=0$ would also not be pivotal.

Let us now consider the joint training of $f$ against an adversary $r$
implemented as a mixture density network modeling $Z|f(X;\theta_f)$ as a mixture
of five gaussians. Similar to $f$, the network architecture of $r$ comprises 2
dense hidden layers of 20 nodes with ReLU activations, but is followed by an
output layer of 15 nodes corresponding to the means, standard deviations and
mixture coefficients of the five gaussians. Output nodes for the mean values
come with linear activations, output nodes for the standard deviations with
exponential activations to ensure positivity, while output nodes for the mixture
coefficients implement the softmax function to ensure positivity and
normalization. When running Algorithm~\ref{alg:adversarial-training} as
initialized with the classifier $f$ obtained previously, adversarial training
effectively reshapes the decision function so it that becomes almost independent
on the nuisance parameter, as shown in Fig.~\ref{fig:toy}. In particular, the
conditional probability densities of the decision scores $f(X;\theta_f)$ are now
very similar to each other, indicating only a small residual  dependency on the
nuisance, as theoretically expected. The dynamics of adversarial training is
illustrated in Fig.~\ref{fig:toy-training}, where the losses ${\cal L}_f$,
${\cal L}_r$ and ${\cal L}_f - \lambda {\cal L}_r$ are evaluated after each
iteration of Algorithm~\ref{alg:adversarial-training}. In the first iterations,
we observe that the global objective ${\cal L}_f - \lambda {\cal L}_r$ is
minimized by making the classifier less accurate, hence the corresponding
increase of ${\cal L}_f$, but which results in a classifier that is more
pivotal, hence the corresponding increase of ${\cal L}_r$ and the total net
benefit. As learning goes, minimizing $E$ then requires making predictions that
are more accurate, hence decreasing ${\cal L}_f$, or that are even less
dependent on $Z$, hence shaping $p_{\theta_r}$ towards the prior $p(z)$. Indeed,
${\cal L}_f$ eventually starts to  decrease, while remaining lower bounded by
$\min_{\theta_f} {\cal L}_f(\theta_f)$ as approximated by the dashed line in the
first plot. Similarly,  ${\cal L}_r$ tends towards the differential entropy
$H(Z)$ of the prior (where $H(Z) = \log(\sigma \sqrt{2 \pi e}) = 1.419$ in the
case of a gaussian with unit variance), as shown by the dashed line in the
second plot. Finally, let us note that the ideal situation of a classifier that
is both optimal and pivotal appears to be unreachable for this problem, as shown
in the third plot by the  offset between ${\cal L}_f - \lambda {\cal L}_r$ and
the dashed line approximating $H({Y|X}) - \lambda H(Z)$.

@@

For several values of $\lambda$, we train a classifier
using  Algorithm~\ref{alg:adversarial-training} but consider the adversarial
term ${\cal L}_r$ conditioned on $Y=0$ only, as outlined in
Sec.~\ref{sec:problem}. Both $f$ and $r$ are neural networks with 3 hidden
layers of 64 nodes and ReLU activations, each terminated by a single final
output node with a sigmoid activation. As in the previous example, adversarial training is initialized
with $f$ pre-trained. Experiments are performed on a subset of
150000 samples for training while AMS is evaluated on an independent test set of
5000000 samples. Both training and testing samples are weighted such that the
null hypothesis corresponded to 1000  of $Y=0$ events and the alternate
hypothesis included an additional 100 $Y=1$ events prior to any thresholding on $f$.
This allows us to probe the efficacy of the
method proposed here in a representative background-dominated high energy physics environment.
Results reported below are averages over 5 runs.

>>>>

Assuming a gaussian prior $z \sim {\cal N}(0,1)$, we start by generating
training data $\{ x_i, y_i, z_i \}_{i=1}^N$, from which we train a neural
network classifier $f$ minimizing ${\cal L}_f(\theta_f)$ without considering its
adversary $r$. The network architecture comprises 2 dense hidden layers of 20
nodes respectively with tanh and ReLU activations, followed  by a dense output layer with a single
node with a sigmoid activation. As shown in Fig.~\ref{fig:toy}, the resulting
classifier is not pivotal, as the conditional probability densities of its
decision scores $f(X;\theta_f)$ show large discrepancies between values $z$ of
the nuisance. While not shown here, a classifier trained only from data
generated at the nominal value $Z=0$ would also not be pivotal.

Let us now consider the joint training of $f$ against an adversary $r$
implemented as a mixture density network modeling $Z|f(X;\theta_f)$ as a mixture
of five gaussians. The network architecture of $r$ comprises 2
dense hidden layers of 20 nodes with ReLU activations, followed by an
output layer of 15 nodes corresponding to the means, standard deviations and
mixture coefficients of the five gaussians. Output nodes for the mean values
come with linear activations, output nodes for the standard deviations with
exponential activations to ensure positivity, while output nodes for the mixture
coefficients implement the softmax function to ensure positivity and
normalization. When running Algorithm~\ref{alg:adversarial-training} as
initialized with the classifier $f$ obtained previously, adversarial training
effectively reshapes the decision function so it that becomes almost independent
on the nuisance parameter, as shown in Fig.~\ref{fig:toy}. In particular, the
conditional probability densities of the decision scores $f(X;\theta_f)$ are now
very similar to each other, indicating only a small residual  dependency on the
nuisance, as theoretically expected. The dynamics of adversarial training is
illustrated in Fig.~\ref{fig:toy-training}, where the losses ${\cal L}_f$,
${\cal L}_r$ and ${\cal L}_f - \lambda {\cal L}_r$ are evaluated after each
iteration of Algorithm~\ref{alg:adversarial-training}. In the first iterations,
we observe that the global objective ${\cal L}_f - \lambda {\cal L}_r$ is
minimized by making the classifier less accurate, hence the corresponding
increase of ${\cal L}_f$, but which results in a classifier that is more
pivotal, hence the corresponding increase of ${\cal L}_r$ and the total net
benefit. As learning goes, minimizing $E$ then requires making predictions that
are more accurate, hence decreasing ${\cal L}_f$, or that are even less
dependent on $Z$, hence shaping $p_{\theta_r}$ towards the prior $p(z)$. Indeed,
${\cal L}_f$ eventually starts to  decrease, while remaining lower bounded by
$\min_{\theta_f} {\cal L}_f(\theta_f)$ as approximated by the dashed line in the
first plot. Similarly,  ${\cal L}_r$ tends towards the differential entropy
$H(Z)$ of the prior (where $H(Z) = \log(\sigma \sqrt{2 \pi e}) = 1.419$ in the
case of a gaussian with unit variance), as shown by the dashed line in the
second plot. Finally, let us note that the ideal situation of a classifier that
is both optimal and pivotal appears to be unreachable for this problem, as shown
in the third plot by the  offset between ${\cal L}_f - \lambda {\cal L}_r$ and
the dashed line approximating $H({Y|X}) - \lambda H(Z)$.

@@

For several values of $\lambda$, we train a classifier
using  Algorithm~\ref{alg:adversarial-training} but consider the adversarial
term ${\cal L}_r$ conditioned on $Y=0$ only, as outlined in
Sec.~\ref{sec:problem}. The architecture of $f$ comprises 3 hidden layers of 64 nodes
respectively with tanh, ReLU and ReLU activations, and is terminated by a single final
output node with a sigmoid activation. The architecture of $r$ is the same,
but uses only ReLU activations in its hidden nodes.
As in the previous example, adversarial training is initialized
with $f$ pre-trained. Experiments are performed on a subset of
150000 samples for training while AMS is evaluated on an independent test set of
5000000 samples. Both training and testing samples are weighted such that the
null hypothesis corresponded to 1000  of $Y=0$ events and the alternate
hypothesis included an additional 100 $Y=1$ events prior to any thresholding on $f$.
This allows us to probe the efficacy of the
method proposed here in a representative background-dominated high energy physics environment.
Results reported below are averages over 5 runs.
